---
title: "p8105_hw2_sjk2254"
author: "Sharon Kulali"
date: "2023-10-02"
output: github_document
---

```{r setup, message = FALSE}
# loading the needed packages

library(tidyverse)
library(readxl)
```


## Problem 2

### Mr. Trash Wheel Sheet

```{r}
# importing mr. trash dataset

mt_df = readxl::read_excel("trash_wheel_data.xlsx", sheet = 1, range = "A2:N550") 
```

```{r}
# cleaning up the dataset

mt_df =
  janitor::clean_names(mt_df) |> 
  drop_na(dumpster)|> 
  mutate(
    year = as.numeric(year),
    vessel_name = "mr trash")
```

```{r}
# updating the homes_powered variable

mt_df = 
  mutate(mt_df, homes_powered = (weight_tons*500)/30)
```

### Professor Trash Wheel Sheet

```{r}
# importing professor trash dataset

pt_df = readxl::read_excel("trash_wheel_data.xlsx", sheet = 2, range = "A2:M97")
```

```{r}
# cleaning up the dataset

pt_df =
  janitor::clean_names(pt_df)|> 
  drop_na(dumpster)|> 
  mutate(vessel_name = "prof trash")
```

```{r}
# updating the homes_powered variable

pt_df = mutate(pt_df, homes_powered = (weight_tons*500)/30)
```

### Gwynnda Trash Wheel Sheet

```{r}
# importing gwynnda trash dataset

gt_df = readxl::read_excel("trash_wheel_data.xlsx", sheet = 4, range = "A2:K110")
```

```{r}
# cleaning up the dataset

gt_df =
  janitor::clean_names(gt_df)|> 
  drop_na(dumpster)|> 
  mutate(vessel_name = "gwyn trash") 
```

```{r}
# updating the homes_powered variable

gt_df = mutate(gt_df, homes_powered = (weight_tons*500)/30)
```

### Combined

```{r}
# creating a new data frame that combines all the data sets

trash_df = bind_rows(mt_df, pt_df, gt_df)
```

### Summary 

Mr. Trash Wheel (`mt_df`), Professor Trash Wheel (`pt_df`), and Gwynnda Trash Wheel (`gt_df`) are water vessels that remove trash from the Jones Falls river in Baltimore, Maryland and dumps them into a dumpster site. The data sets `mt_df`, `pt_df`, and `gt_df` include information about the dumpster number, date of collection, amount of total litter, and the specific litter type. The total weight of trash collected by Professor Trash Wheel was `r sum(pull(pt_df, weight_tons))` tons. The total number of cigarette butts collected by Gwynnda Trash Wheel was `r sum(filter(gt_df, month == "July" & year == 2021)$cigarette_butts)` tons. In the combined data set of all three trash vessels (`trash_df`), there are `r nrow(trash_df)` observations with an average litter weight of `r mean(pull(trash_df, weight_tons))` tons and information collected from `r min(pull(trash_df, year))` to `r max(pull(trash_df, year))`. Notably, the plastic bags variable had `r sum(is.na(pull(trash_df, plastic_bags)))` missing values.




